---
title: "Data Import"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

Let's import a dataset. 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv") #Don't use read.csv
```

Look at your data!

```{r}
names(litters_df)
```
Update the names in 'litters_df'. 

```{r}
litters_df = 
  janitor::clean_names(litters_df) #Just want clean_names function from janitor package without loading the whole package
names(litters_df)
litters_df
```

Sometimes skimming data is neat?

```{r}
skimr::skim(litters_df)
```

## Fix the missingness (using two hashtags to make font of this title bigger)

```{r}
litters_df =
    read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", "")) #This will change the "." to NA to keep consistency. 
```

## Import the FAS_pups dataset. 

```{r}
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", 
          skip = 3, #Skip first three rows since empty 
          na = c("NA", ".", ""))

#Clean names
pups_df = 
  janitor::clean_names(pups_df)
names(pups_df)
```

## Okay what about Excel 

CSVs are really great but sometimes you get an excel file. Try to use CSVs if possible though...

```{r}
mlb_df <-  
  read_excel("data_import_examples/mlb11.xlsx")
mlb_df
```

Now what about when the tables aren't neatly displayed in Excel and have formatting. 

Import LOR word counts. 

```{r}
fotr_df <- read_excel("data_import_examples/LotR_Words.xlsx", range = "B3:D6") #Specifies the range of which of the three tables we want from the Excel. 
```

## SAS????

Import the PULSE data. 

```{r}
pulse_df <- read_sas("data_import_examples/public_pulse_data.sas7bdat")

pulse_df <- janitor::clean_names(pulse_df)
```



##Why do I hate read.csv so much?? It prints worse and doesn't organize it as readable.

```{r, eval=FALSE}
litters_df_base <- 
    read.csv("data_import_examples/FAS_litters.csv")
```

